Imputation of Incomplete Multilevel Data with \proglang{R}

### 2023-11-21 Manuscript/Imputation_of_Incomplete_Multilevel_Data.Rmd
move to quarto

Many datasets include individuals that are clustered together, for example in geographic regions, or even different studies. In the simplest case, individuals (e.g., students) are nested within a single cluster (e.g., school classes). More complex clustered structures may occur when there are multiple hierarchical levels (e.g., students in different schools or patients within hospitals within regions across countries), or when the clustering is non-nested (e.g., electronic health record data from diverse settings and populations within large databases). With clustered data we generally assume that individuals from the same cluster tend to be more similar than individuals from other clusters. In statistical terms, this implies that observations from the same cluster are not independent and may in fact be correlated. If this correlation is left unaddressed, estimates of *p* values, confidence intervals even model parameters are prone to bias  [@loca01]. Statistical methods for clustered data typically adopt hierarchical models that explicitly describe the grouping of observations. These models are also known as 'multilevel models', 'hierarchical models', 'mixed effect models', 'random effect models', and in the context of time-to-event data as 'frailty models'. Table \ref{tab:clus} provides an overview of some key concepts in multilevel modeling.

>>>>

Understanding and effectively handling missing data within multilevel structures are crucial for preserving the integrity of analyses and deriving accurate inferences.

Many datasets include individuals that are clustered together, for example by geographic region. In the simplest case, each individual is a unit (e.g., a student) which is nested within a single cluster (e.g., a school class). More complex clustered structures may occur when there are multiple hierarchical levels (e.g., patients within hospitals within regions across countries), or when the clustering is non-nested (e.g., electronic health record data from diverse settings and populations within large databases). The hierarchical structures of such clustered data may be captured using multilevel analysis techniques.

With clustered data we generally assume that individuals from the same cluster tend to be more similar than individuals from other clusters. In statistical terms, this implies that observations from the same cluster are not independent and may in fact be correlated. If this correlation is left unaddressed, estimates of *p* values, confidence intervals even model parameters are prone to bias  [@loca01]. Statistical methods for clustered data typically adopt hierarchical models that explicitly describe the grouping of observations. These models are also known as 'multilevel models', 'hierarchical models', 'mixed effect models', 'random effect models', and in the context of time-to-event data as 'frailty models'. Table \ref{tab:clus} provides an overview of some key concepts in multilevel modeling.

In the context of multilevel data, missingness can occur at different levels, complicating the interpretation and reliability of results.

### 2023-11-21 Manuscript/Imputation_of_Incomplete_Multilevel_Data.tex
move to quarto

\usepackage{graphicx}
\usepackage{mathtools}
\usepackage{ulem}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> #ggmice::plot_pattern(data=Obesity,cluster="Cluster") does not work!! bug
R> #ggmice::plot_pattern(data=Obesity,cluster=Obesity$Cluster) does not work!! bug
R> library(ggpubr)
R> myplots <- lapply(1:5, function(i) {
+   ggmice::plot_pattern(setDT(Obesity)[Cluster==i])+
+   ggplot2::ggtitle(paste0("Cluster", i))
+  })
R> ggarrange(myplots[[1]], myplots[[3]], nrow=1,common.legend = TRUE, legend="bottom")
\end{CodeInput}
\begin{figure}

{\centering \includegraphics[width=0.7\linewidth]{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/unnamed-chunk-12-1}

\caption[Missing pattern]{Missing pattern}\label{fig:unnamed-chunk-12}
\end{figure}
\end{CodeChunk}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> Nulmodel <- lme4::lmer(BMI ~ 1 + (1|Cluster), data = Obesity)
R> performance::icc(Nulmodel)
\end{CodeInput}
\begin{CodeOutput}
# Intraclass Correlation Coefficient

    Adjusted ICC: 0.362
  Unadjusted ICC: 0.362
\end{CodeOutput}
\end{CodeChunk}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> library(micemd)
R> meth_mar <- micemd::find.defaultMethod(Obesity, ind.clust=1, I.small = 7,
+                                ni.small = 100, prop.small = 0.4)
R> meth_mar["BMI"]<- "~ I(Weight / (Height)^2)"
R> meth_mar["Age"]<-"2l.2stage.pmm" 
\end{CodeInput}
\end{CodeChunk}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> pred_mar <- mice(Obesity, maxit = 0)$pred 
R> pred_mar[,"Cluster"] <- -2 # clustering variable
R> pred_mar[,"Time"] <- 0
R> pred_mar[pred_mar==1] <- 2
R> pred_mar[c("Height", "Weight"), "BMI"] <- 0
R> ggmice::plot_pred(pred_mar)
\end{CodeInput}

\begin{center}\includegraphics[width=0.7\linewidth]{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/obesity-predmar-1} \end{center}

\begin{CodeInput}
R> imp_mar <- mice::mice(data = Obesity, meth = meth_mar, pred = pred_mar,
+                 m=10, seed = 123, printFlag = FALSE)
\end{CodeInput}
\end{CodeChunk}

\begin{CodeChunk}
\begin{CodeInput}
R> summary(complete(imp_mar,"long")$Weight)
\end{CodeInput}
\begin{CodeOutput}
   Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
 -19.48   68.45   81.65   81.28   94.12  160.76 
\end{CodeOutput}
\end{CodeChunk}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> summary(complete(imp_mar_pmm,"long")$Weight)
\end{CodeInput}
\begin{CodeOutput}
   Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
  28.35   69.25   82.43   82.41   94.76  134.61 
\end{CodeOutput}
\begin{CodeInput}
R> ggmice::plot_trace(imp_mar_pmm, "Weight")
\end{CodeInput}

\begin{center}\includegraphics{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/obesity-predmar_pmm1-1} \end{center}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> pred_mnar <- pred_mar
R> pred_mnar["Weight","Time"]<- -3
R> ggmice::plot_pred(pred_mnar)
\end{CodeInput}

\begin{center}\includegraphics{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/obesity-predmnar-1} \end{center}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> imp_mnar<- mice(data = Obesity, meth = meth_mnar, pred = pred_mnar,
+                 m=10, seed = 123, printFlag = FALSE)
R> summary(complete(imp_mnar,"long")$Weight)
\end{CodeInput}
\begin{CodeOutput}
   Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
  14.09   71.68   85.33   85.79   99.31  186.67 
\end{CodeOutput}
\end{CodeChunk}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> summary(complete(imp_mnar,"long")$Weight)
\end{CodeInput}
\begin{CodeOutput}
   Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
  14.09   71.68   85.33   85.79   99.31  186.67 
\end{CodeOutput}
\end{CodeChunk}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> summary(complete(imp_mnar_pmm,"long")$Weight)
\end{CodeInput}
\begin{CodeOutput}
   Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
  28.35   71.51   85.09   85.64   99.00  134.61 
\end{CodeOutput}
\begin{CodeInput}
R> ggmice::plot_trace(imp_mar_pmm, "Weight")
\end{CodeInput}

\begin{center}\includegraphics{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/obesity-predmnarp1-1} \end{center}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> library(ggplot2)
R> cc_rs<- with(setDT(Obesity)[complete.cases(Obesity),],
+              lme( BMI ~ Age + FamOb + Gender, random=~1+Age|Cluster))
R> mar_rs <- with(imp_mar,lme( BMI ~ Age + FamOb + Gender,random=~1+Age|Cluster))
R> mar_pmm_rs <- with(imp_mar_pmm,lme( BMI ~ Age + FamOb + Gender,random=~1+Age|Cluster))
R> mnar_rs<- with(imp_mnar,lme(BMI ~ Age + FamOb + Gender,random=~1+Age|Cluster))
R> mnar_pmm_rs<- with(imp_mnar_pmm, lme(BMI ~ Age + FamOb + Gender,random=~1+Age|Cluster))
R> list_models<-list(cc_rs,mar_rs,mar_pmm_rs,mnar_rs,mnar_pmm_rs)
R> plot_models(list_models,
+             mod_name = c("Complete case", "MAR","MAR_pmm", "MNAR", "MNAR_pmm"))
\end{CodeInput}

\begin{center}\includegraphics{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/models-1} \end{center}

@@

@@

\hypertarget{conclusion}{%
\section{Conclusion}\label{conclusion}}

>>>>

\usepackage{graphicx}
\usepackage{mathtools}
\usepackage{ulem}
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> library(ggpubr)
R> myplots <- lapply(1:5, function(i) {
+   ggmice::plot_pattern(setDT(Obesity)[Cluster==i])+
+   ggplot2::ggtitle(paste0("Cluster", i))
+  })
R> ggarrange(myplots[[1]], myplots[[3]], nrow=1,common.legend = TRUE, legend="bottom")
\end{CodeInput}
\end{CodeChunk}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> Nulmodel <- lme4::lmer(BMI ~ 1 + (1|Cluster), data = Obesity)
R> #Performance::ICC(Nulmodel)
\end{CodeInput}
\end{CodeChunk}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> meth_mar <- find.defaultMethod(Obesity, ind.clust=1, I.small = 7,
+                                ni.small = 100, prop.small = 0.4)
R> meth_mar["BMI"]<- "~ I(Weight / (Height)^2)"
R> meth_mar["Age"]<-"2l.2stage.pmm" 
\end{CodeInput}
\end{CodeChunk}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> pred_mar <- quickpred(Obesity)   # predictor matrix
R> pred_mar[,"Cluster"] <- -2 # clustering variable
R> pred_mar[,"Time"] <- 0
R> pred_mar[pred_mar==1] <- 2
R> pred_mar[c("Height", "Weight"), "BMI"] <- 0
R> ggmice::plot_pred(pred_mar)
R> imp_mar <- mice(data = Obesity, meth = meth_mar, pred = pred_mar,
+                 m=10, seed = 123, printFlag = FALSE)
\end{CodeInput}
\end{CodeChunk}

\begin{CodeChunk}
\begin{CodeInput}
R> summary(complete(imp_mar,"long")$Weight)
\end{CodeInput}
\end{CodeChunk}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> summary(complete(imp_mar_pmm,"long")$Weight)
R> #ggmice::ggplot_trace(imp_mar_pmm, "Weight")
\end{CodeInput}
\end{CodeChunk}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> pred_mnar <- pred_mar
R> pred_mnar["Weight","Time"]<- -3
R> #ggmice::plot_pred(pred_mnar)
\end{CodeInput}
\end{CodeChunk}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> imp_mnar<- mice(data = Obesity, meth = meth_mnar, pred = pred_mnar,
+                 m=10, seed = 123, printFlag = FALSE)
R> summary(complete(imp_mnar,"long")$Weight)
\end{CodeInput}
\end{CodeChunk}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> summary(complete(imp_mnar,"long")$Weight)
\end{CodeInput}
\end{CodeChunk}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> summary(complete(imp_mnar_pmm,"long")$Weight)
R> #ggmice::ggplot_trace(imp_mar_pmm, "Weight")
\end{CodeInput}
\end{CodeChunk}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> library(ggplot2)
R> cc_rs<- with(setDT(Obesity)[complete.cases(Obesity),],
+              lme( BMI ~ Age + FamOb + Gender, random=~1+Age|Cluster))
R> mar_rs <- with(imp_mar,lme( BMI ~ Age + FamOb + Gender,random=~1+Age|Cluster))
R> mar_pmm_rs <- with(imp_mar_pmm,lme( BMI ~ Age + FamOb + Gender,random=~1+Age|Cluster))
R> mnar_rs<- with(imp_mnar,lme(BMI ~ Age + FamOb + Gender,random=~1+Age|Cluster))
R> mnar_pmm_rs<- with(imp_mnar_pmm, lme(BMI ~ Age + FamOb + Gender,random=~1+Age|Cluster))
R> list_models<-list(cc_rs,mar_rs,mar_pmm_rs,mnar_rs,mnar_pmm_rs)
R> plot_models(list_models,
+             mod_name = c("Complete case", "MAR","MAR_pmm", "MNAR", "MNAR_pmm"))
\end{CodeInput}
\end{CodeChunk}

@@

Data are simulated and included in the \texttt{micemd} package. We will
use the following variables:

\begin{itemize}
\tightlist
\item
  \texttt{region} Cluster variable,
\item
  \texttt{gender} Gender (0=male, 1=female),
\item
  \texttt{age} Age of the patient,
\item
  \texttt{height} Height in meters,
\item
  \texttt{weight} Weight in kilograms,
\item
  `time`` Response time in minutes (inclusion-restriction variable).
\end{itemize}

Load the data:

\begin{CodeChunk}
\begin{CodeInput}
R> data("data_heckman", package = "micemd")
\end{CodeInput}
\end{CodeChunk}

Load the data:

\begin{CodeChunk}
\begin{CodeInput}
R> #data("obesity", package = "miceheckman")
\end{CodeInput}
\end{CodeChunk}

Select relevant variables:

\begin{CodeChunk}
\begin{CodeInput}
R> #dat <- select(obesity, -"bmi")
\end{CodeInput}
\end{CodeChunk}

Create predictor matrix:

\begin{CodeChunk}
\begin{CodeInput}
R> #pred <- quickpred(dat)   # predictor matrix
R> #pred["weight","cluster"]<- -2 # clustering variable
R> #pred["weight","rt"] <- -3 #inclusion-restriction variable
R> #plot_pred(pred)
\end{CodeInput}
\end{CodeChunk}

Set the Heckman model as imputation method:

\begin{CodeChunk}
\begin{CodeInput}
R> #meth <- make.method(dat) # methods vector
R> #meth["weight"]<-"2l.heckman"
\end{CodeInput}
\end{CodeChunk}

\begin{CodeChunk}
\begin{CodeInput}
R> plot_pattern(dat)
\end{CodeInput}

\begin{center}\includegraphics{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/obesity-md-1} \end{center}

Create predictor matrix:

\begin{CodeChunk}
\begin{CodeInput}
R> pred <- quickpred(dat)   # predictor matrix
R> pred[,"Cluster"] <- -2 # clustering variable
R> pred[, "Weight"] <- -3 #inclusion-restriction variable
R> plot_pred(pred)
\end{CodeInput}
\end{CodeChunk}

Set the Heckman model as imputation method:

\begin{CodeChunk}
\begin{CodeInput}
R> meth <- make.method(dat) # methods vector
R> meth["weight"]<-"2l.heckman"
\end{CodeInput}
\end{CodeChunk}

Impute the missingness:

\begin{CodeChunk}
\begin{CodeInput}
R> imp <- mice(
+   dat, # dataset with missing values
+   m = 5, # number of imputations
+   maxit = 10,
+   meth = meth, #imputation method vector
+   pred = pred, #imputation predictors matrix
+   meta_method = "reml",
+   printFlag = FALSE
+ )
\end{CodeInput}
\end{CodeChunk}

\hypertarget{discussion}{%
\section{Discussion}\label{discussion}}

@@
